Green Nest Material Meta-Analysis

### 2025-06-17 manuscript_GNM.qmd
Some sensitivity analysis added

>>>>

# Introduction

Nest construction is a fundamental aspect of the bird breeding cycle, providing structural support and protection for eggs, parents, and nestlings against predators and environmental pressures, and in some species, even playing a role in sexual display during mating [@mainwaring2014; @hansell2000]. Nest building behaviour is increasingly recognized as an expression of extended phenotype, demonstrating interspecific and intraspecific trait variation [@moreno2012; @perez2023]. While most birds use organic plant materials- twigs, branches, grass, and moss- some incorporate inorganic items such as stones, feathers, snake skin and even anthropogenic materials [@sergio2011; @sheard2024]. These non-structural additions suggest that nests are more than just physical shelters; they can be complex tools shaped by both natural and sexual selection for multiple functions.

An intriguing nesting behaviour observed in some birds is the use of fresh green plant material (green nest material or GNM). This includes leaves, sprigs, herbs and branches added during or even after primary nest construction has completed. Importantly, these green nest materials differ from the structural components, being selected from a small, non-random subset of local flora [@clark1985; @garrido-bautista2023], often characterized by high levels of aromatic volatile compounds (commonly used by Passerines, [@mennerat2009]) and secondary metabolites such as oils and resins (preferred by raptors, [@ontiveros2007]). The costs associated with gathering these materials beyond the territory and maintaining fresh greenery through the breeding season imply adaptive significance of the GNM.

Several hypotheses have been proposed to explain the adaptive function of green nest material (GNM). The courtship hypothesis suggests that males use aromatic plants as olfactory signals to attract mates or advertise nest site occupancy, particularly in cavity-nesting species where visual cues are limited. This idea traces back to the early natural history observations, beginning with @kessel1957, who noted that male European Starlings (*Sturnus vulgaris*) carried fresh plant material to nests while singing and displaying to females. Subsequent studies in European and Spotless Starlings (*Sturnus unicolor*) support this view: unpaired males conspicuously deliver greenery during courtship but stop after egg-laying begins [@brouwer2004; @veiga2006], and they increase this behaviour in the presence of potential mates \[[@eens1993]\]. Females often prefer nests with fresh herbs, suggesting that GNM may serve as a sexually selected signal of male quality and potentially increase fitness benefits for females. The courtship hypothesis has been also extended to species like Blue Tits (*Cyanistes caeruleus*), where it has been proposed that GNM might serve as a signal from females to the males [@tomás2013]. However, in both starlings and blue tits, the link between plant-bringing material in males and the fitness benefits for females still remain unclear.

Another hypothesis to explain the adaptive benefit of GNM is the widely tested nest protection hypothesis (NPH), which proposes that volatile compounds in green material reduce parasite abundance or pathogen loads, indirectly enhancing offspring survival. In an influential early comparitive study in raptors (Falconiformes), @wimberger1984 found indirect evidence for the reduction of ectoparasites in the nests with green nest material. @clark1988 later provided experimental support in European Starlings, showing that preferred herbs contained insect-repellent and antimicrobial volatile. They also outlined key criteria for evaluation of the NPH, including selective plant use, bioactive chemical properties and demonstrable adaptive benefits. Field observation and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@dawson2004; @lafuma2001; @mennerat2009b; @quiroga2012] have shown mixed results, and in some cases, GNM improved nestling condition without reducing parasite abundance [@gwinner2000]. These findings led to the development of drug hypothesis, which suggests that the aromatic GNM directly enhance nestling development or immune function, independent of the parasite control [@gwinner2000; @mennerat2009a]. Thus, while both hypotheses posit adaptive benefits of GNM, the drug hypothesis emphasizes a pharmacological mechanism, and the two are not mutually exclusive. Several other adaptive functions of GNM have also been proposed. These functions do not refer to the aromatic properties of GM, but could still provide some benefits such as enhance the crypticity of the nest [@skutch1957], provide insulation for eggs and nestling [@mertens1977], provide protection from sun, signal nest occupancy or simply fill the gaps between nesting materials [@lyons1987]. These alternative functional roles have received little support or attention.

Despite several observational and experimental studies and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], a comprehensive quantitative synthesis across studies and hypotheses has been lacking. This has limited our understanding of the generality and magnitude of GNM effects on fitness benefits. To address this gap, we conducted a systematic review and meta-analysis to assess the overall effect of GNM on various fitness proxies. We explicitly evaluate the evidence for the courtship and parental-care- the latter combining the nest protection and drug hypotheses due to their non-exclusivity. The drug hypothesis can only be supported by improved nestling health while ensuring that GNM has no protective effect against any parasite or pathogen.

In addition to testing these primary hypotheses, we further explore the context-dependency in the research field and examine the potential sources of variation in the results (heterogeneity) in the GNM effects. Specifically, we explored whether the effect of GNM varied depending on the type of parasite (arthropods vs. microorganisms), the timing of GNM addition (e.g., before hatching, after hatching, or continuously), the experimental design (e.g., aromatic vs. non-aromatic vs. no added material), the focal species, and the type of fitness trait measured (e.g., physiology, morphology, reproduction). These exploratory moderators could allow us to assess context dependence and identify ecological or methodological patterns that may explain inconsistencies in previous findings. Together, our evidence synthesis approach provides the first quantitative test of the adaptive value of green nest material across bird species and research contexts.

### 2025-06-17 notebook/data_analysis.qmd
Some sensitivity analysis added

@@

```{r}
# This analysis only includes dataset that has a proxy relating to comaparision type
dataset_lnRR_design <- dataset_lnRR %>%
  filter(!is.na(comparision_type))%>%
  mutate(comparision_type= as.factor(comparision_type))

VCV_lnRR_design <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_design,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
lnRR_design <-
  metafor::rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_design ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + comparision_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_lnRR_design )

@@

# This analysis only includes dataset that has a proxy relating to comaparision type
dataset_SMDH_design <- dataset_SMDH %>%
  filter(!is.na(comparision_type))%>%
  mutate(comparision_type= as.factor(comparision_type))

VCV_SMDH_design <- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH_design,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
SMDH_design <-
  metafor::rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_design ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + comparision_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_SMDH_design )

@@

>>>>

### Sensitivity Analysis excluding the parasite and pathogen load fitness proxies

## For the lnRR effect size...
dataset_lnRR_no_PCH<-dataset_lnRR%>%filter(Hypothesis!="PCH")

## The dataset is reduced to almost half
nrow(dataset_lnRR)-nrow(dataset_lnRR_no_PCH)

VCV_lnRR_no_PCH <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_no_PCH,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SA_lnRR_no_PCH<- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_no_PCH, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_lnRR_no_PCH)

## For the SMDH effect size..

dataset_SMDH_no_PCH<-dataset_SMDH%>%filter(Hypothesis!="PCH")

## The dataset is reduced to almost half
nrow(dataset_SMDH)-nrow(dataset_SMDH_no_PCH)

VCV_SMDH_no_PCH <- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH_no_PCH,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SA_SMDH_no_PCH<- rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_no_PCH, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_SMDH_no_PCH)
```

### Sensitivity Analysis for excluding non-aromatic and blank (the two control group comparison)

## For the lnRR effect size...
dataset_lnRR_no_comp3<-dataset_lnRR%>%filter(comparision_type!=3)
## Only some rows are removed... 
nrow(dataset_lnRR)-nrow(dataset_lnRR_no_comp3)

VCV_lnRR_no_comp3 <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_no_comp3,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SA_lnRR_no_PCH<- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_no_PCH, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_lnRR_no_PCH)
```

@@

```{r}
# This analysis only includes dataset that has a proxy relating to comaparision type
dataset_lnRR_design <- dataset_lnRR %>%
  mutate(comparision_type= case_when(is.na(comparision_type) ~ 4, TRUE ~ comparision_type))

VCV_lnRR_design <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_design,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
lnRR_design <-
  metafor::rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_design ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + as.factor(comparision_type),
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_lnRR_design )

@@

# This analysis only includes dataset that has a proxy relating to comaparision type
dataset_SMDH_design <- dataset_SMDH %>%
  mutate(comparision_type= case_when(is.na(comparision_type) ~ 4, TRUE ~ comparision_type))

VCV_SMDH_design <- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH_design,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
SMDH_design <-
  metafor::rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_design ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + as.factor(comparision_type),
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_SMDH_design )

@@

## All-in Model for heterogeneity

```{r}
## For lnRR complete dataset

## I am changing the year of unpublished studies to 2025 which is to now, assuming that if they were to be published now, what would be the effect. I will keep in mind that in this model if I do find an evidence of publication bias over years, I should keep in mind it is non-independent of the unpublished studies

dataset_lnRR_all_in<-dataset_lnRR%>%mutate(year_publication = case_when(is.na(year_publication) ~ 2025, TRUE ~ year_publication))%>%
  mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])%>%
  mutate(effective_n_inv=1/((4*effective_n_experiment*effective_n_control)/(effective_n_experiment+effective_n_control)))%>%
  mutate(sqrt_effective_n_inv=sqrt(effective_n_inv))%>%
  mutate(comparision_type= case_when(is.na(comparision_type) ~ 4, TRUE ~ comparision_type))

VCV_lnRR_all_in<- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_all_in,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

lnRR_all_in <- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_all_in, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ time_of_gnm_addition + 
                      year_MeanCenter + 
                      sqrt_effective_n_inv + 
                      as.factor(comparision_type) + 
                      bird_species + 
                      trait_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_lnRR_all_in)

## For SMDH complete published dataset

dataset_SMDH_all_in<-dataset_SMDH%>%mutate(year_publication = case_when(is.na(year_publication) ~ 2025, TRUE ~ year_publication))%>%
  mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])%>%
  mutate(effective_n_inv=1/((4*effective_n_experiment*effective_n_control)/(effective_n_experiment+effective_n_control)))%>%
  mutate(sqrt_effective_n_inv=sqrt(effective_n_inv))%>%
  mutate(comparision_type= case_when(is.na(comparision_type) ~ 4, TRUE ~ comparision_type))

VCV_SMDH_all_in<- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH_all_in,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SMDH_all_in<- rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_all_in, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ time_of_gnm_addition + 
                      year_MeanCenter + 
                      sqrt_effective_n_inv + 
                      as.factor(comparision_type) + 
                      bird_species + 
                      trait_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_SMDH_all_in)
```
